Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-04 paper_jasa.tex
working on experimental explanation section

@@

The experiment is designed to study the ability of human observers to detect the effect of a single variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated a range of values of $\beta_2~ (=0, 1, 3, 5, 7, 8, 10, 16)$, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical UMP test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson(30) distribution. \green{Explain why this was used.} \red{There was not any real reason for this. It was just arbitrarily chosen to generate some data. When I first started to think of applying visual inference in regression setting, I was thinking of a dataset (I found in a book) of weight of babies by their gestational age (in week) where author wanted to test the difference of mean weight for gender.   Thus for our experiment I generated values similar to a variable like age. the discrete variable was thought similar to gender.} Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets''.

{\bf Steps to obtain lineup}
\begin{enumerate}
	\item For a given parameter setting obtain actual data set
	\item Fit the null model to the actual data and obtain residuals and other parameter estimates ($\hat{\beta}$ and $\hat{\sigma}$)
	\item Generate actual data plot using the procedure as described in table \ref{tbl:stat_multiple}
	\item Generate 19 null (residual) data sets from $N(0, {\hat{\sigma}}^2$) and obtain 19 null plots
	\item Finally randomly place 19 null plots and one actual plot in a layout (4X5) to get a lineup 
\end{enumerate}

 Figure \ref{fig:test_category} is an example of one of these lineups. The visual test statistic is a side-by-side boxplot. The actual data plot is number 13 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.

{\bf Number of evaluations} required for each lineup is determined based on the precision level of the estimated power. Suppose $\gamma$ denotes the classical UMP test power for each parameter combination shown in table \ref{tbl:experiment_params}. Since expected power of visual inference is very close to the power of classical UMP test as in figure \ref{fig:power_expected} the proportion of successful evaluations of lineups should give a rough estimate for  $\gamma$. For a given proportion $\gamma$ we want to have margin of error (ME) to be 0.1. Thus we have $ME =1.96 \sqrt{ \frac 1 n_{\gamma} \gamma(1-\gamma)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$  
Each subject viewed at least 10 lineups and we allowed them to provide more evaluations if they wanted. Depending on the parameter combinations we group the lineups in different difficulty levels such as easy, medium, hard and mixed as shown in table \ref{tbl:dist_lineup1}. For each difficulty level a specific number of lineups were randomly picked to show for evaluation. This number is chosen so that total number of evaluations for each lineup for that group exceed the threshold $n_{\gamma}$. To satisfy this plan we needed to recruit at least 300 subjects.

\green{Did you in fact bin like this???}\red{For each parameter combination a set of 1000 actual data are generated and their associated p-values were computed. This produces an empirical distribution of p-value under alternative hypothesis. Finally, three data sets that have 16th, 50th and 84th quantile of the p-values were taken. Similar procedure was used to generate 19 null plots based on the empirical distribution of p-value under null hypothesis}. For added control, to ensure a signal in the simulated actual data a blocking structure was used to filter data sets. A 1000 sets were generated for each parameter combination and the traditional $t$-statistic and $p$-value associated with $H_0: \beta_2=0$ were calculated. The 3 replicates were drawn from each of three blocks of $p$-values: (0.0-$q_{33}$), ($q_{33}$-$q_{66}$), ($q_{66}$-1) where $q_i$ is the $i$th percentile in the distribution of the $p$-values. Additional control was applied to the 19 null plots. Because the distribution of these $p$-values should follow a Uniform(0,1) distribution, data sets were binned on this range by $p$-value, and a data set was randomly selected from each bin.

@@

@@

%\end{multicols}

>>>>

After many small pilot studies with local personnel, it was clear that some care was needed to set up the human subjects experiments. It was best for a subject to see a block of 10 lineups with varying difficulty, with a reasonable number of ``easy'' lineups. The explanations about each experiment (below) includes and explanation of how the lineups were sampled and provided to the observers.

@@

The experiment is designed to study the ability of human observers to detect the effect of a single variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated a range of values of $\beta_2~ (=0, 1, 3, 5, 7, 8, 10, 16)$, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical UMP test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. %\green{Explain why this was used.} \red{There was not any real reason for this. It was just arbitrarily chosen to generate some data. When I first started to think of applying visual inference in regression setting, I was thinking of a dataset (I found in a book) of weight of babies by their gestational age (in week) where author wanted to test the difference of mean weight for gender.   Thus for our experiment I generated values similar to a variable like age. the discrete variable was thought similar to gender.}
Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual ydata plot was generated by making a side-by-side boxplot of the residuals (Table\ref{tbl:stat_multiple} case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2$, and these were plotted in the same way as side-by-side boxplots. The actual data plot was randomly placed among these null data plots to produce the lineup. Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 13 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.  \green{I think it is plot11, not 13!!!!!!!! Please check!}

%{\bf Steps to obtain lineup}
%\begin{enumerate}
%	\item For a given parameter setting obtain actual data set
%	\item Fit the null model to the actual data and obtain residuals and other parameter estimates ($\hat{\beta}$ and $\hat{\sigma}$)
%	\item Generate actual data plot using the procedure as described in table \ref{tbl:stat_multiple}
%	\item Generate 19 null (residual) data sets from $N(0, {\hat{\sigma}}^2$) and obtain 19 null plots
%	\item Finally randomly place 19 null plots and one actual plot in a layout (4X5) to get a lineup 
%\end{enumerate}

 %Figure \ref{fig:test_category} is an example of one of these lineups. The visual test statistic is a side-by-side boxplot. The actual data plot is number 13 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.

The number of evaluations required for each lineup to provide reasonable estimates of the power is determined by the variance of the estimated power. Suppose $\gamma$ denotes the classical UMP test power for each parameter combination shown in table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of classical UMP test (Figure \ref{fig:power_expected}) the proportion of successful evaluations of lineups should give a rough estimate for  $\gamma$. For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.1. Thus we have $ME =1.96 \sqrt{ \frac 1 n_{\gamma} \gamma(1-\gamma)} \le 0.05$ \green{0.05 or 0.1???} which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$

Each subject viewed at least 10 lineups and we allowed them to provide more evaluations if they wanted. Depending on the parameter combinations we group the lineups in different difficulty levels such as easy, medium, hard and mixed (actual numbers are given in the Appendix). For each difficulty level a specific number of lineups were randomly picked to show for evaluation. This number is chosen so that total number of evaluations for each lineup for that group exceed the threshold $n_{\gamma}$. To satisfy this plan we needed to recruit at least 300 subjects.

@@

%\green{Did you in fact bin like this???}\red{For each parameter combination a set of 1000 actual data are generated and their associated p-values were computed. This produces an empirical distribution of p-value under alternative hypothesis. Finally, three data sets that have 16th, 50th and 84th quantile of the p-values were taken. Similar procedure was used to generate 19 null plots based on the empirical distribution of p-value under null hypothesis}. For added control, to ensure a signal in the simulated actual data a blocking structure was used to filter data sets. A 1000 sets were generated for each parameter combination and the traditional $t$-statistic and $p$-value associated with $H_0: \beta_2=0$ were calculated. The 3 replicates were drawn from each of three blocks of $p$-values: (0.0-$q_{33}$), ($q_{33}$-$q_{66}$), ($q_{66}$-1) where $q_i$ is the $i$th percentile in the distribution of the $p$-values. Additional control was applied to the 19 null plots. Because the distribution of these $p$-values should follow a Uniform(0,1) distribution, data sets were binned on this range by $p$-value, and a data set was randomly selected from each bin.

@@

%\end{multicols}
\section*{Appendix}
